prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-04 inst/MyArticle/paper.Rmd
updates to abstract and to the introduction.

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Paolo Nanni$^{\star}$
  - Jonas Grossmann$^{\star}$
  - Maria d'Errico$^{\star}$
  - Ralph Schlapbach$^{\star}$
  - Christian Panse$^{\star}$
  - Witold E. Wolski $^{\star}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich; Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. Correspondence$^\star$\texttt{wew@fgcz.ethz.ch}]
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programmer interface in R that transparently supports a variety of well principled statistical inference procedures.
    The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates essential steps of the mass spectrometry-based label-free quantification data analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors using various statistical models. It also can be used to develop applications that produce interactive data analysis reports in PDF or HTML format. 
     Furthermore, the package can be used to benchmark data acquisition, data preprocessing, or data modeling methods using gold standard datasets. Using the benchmark functionality, we show that the methods implemented allow performing sensitive and specific differential expression analysis. 
     The R package is available through https://github.com/fgcz/prolfqua.

@@

Proteins are carriers of function and structure in living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. However, identifying and quantifying intact proteins in complex samples containing hundreds of proteins is not possible using existing methods at the moment. Hence, the bottom-up mass spectrometric method is employed [@Bubis2017, @da2020philosopher]. Proteins that are extracted from biological samples are specifically and reproducible digested into protein fragments - peptides. The peptides are first separated by their chemical and physical properties using High-Pressure Liquid Chromatography (HPLC). Afterward, they are weighted, identified, and quantified using mass spectrometric techniques, e.g., Electro-Spray-Ionization Mass Spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences[@eng2015deeper, @yu2016pipi, @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast, @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath, @demichev2020dia] are counted and aggregated. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

Protein quantification proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with increasing numbers of samples are studid, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a specific factors and their interactions are significant. Furthermore it can be tested if differences between factor levels can be estimated, and their significance tested. Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [@subramanian2005gene] or validation experiments.

Several packages exist to model mass spectrometry LFQ experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features. For example, `r BiocStyle::Biocpkg("MSstats")` infers the model from the sample annotation and generates the model formula. At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level fold change and $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage known from `r BiocStyle::Biocpkg("limma")`. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

When analyzing parallel-group designs using a single explanatory variable, we can use all R packages; however, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that packages such as $limma$ and $proDA$ allow us to a comprehensive variety of hypotheses; however, in-depth knowledge of design matrix specification is required [@law2020guide].

>>>>

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Paolo Nanni$^{\star}$
  - Jonas Grossmann$^{\star}$
  - Maria d'Errico$^{\star}$
  - Ralph Schlapbach$^{\star}$
  - Christian Panse$^{\star}$
  - Witold E. Wolski $^{\star}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich; Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. Correspondence$^\star$\texttt{wew@fgcz.ethz.ch}]
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programmer interface in R that transparently supports a variety of well principled statistical procedures. The package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing.
    The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates, using an R6 application programmer interface, essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The R package is available through https://github.com/fgcz/prolfqua.

@@

Proteins are carriers of function and structure in living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. However, identifying and quantifying intact proteins in complex samples containing hundreds of proteins is not possible using existing methods at the moment. Hence, the bottom-up mass spectrometric method is employed [@Bubis2017, @da2020philosopher]. Proteins that are extracted from biological samples are specifically and reproducible digested into protein fragments - peptides. The peptides are first separated by their chemical and physical properties using High-Pressure Liquid Chromatography (HPLC). Afterward, they are weighted, identified, and quantified using mass spectrometric techniques, e.g., Electro-Spray-Ionization Mass Spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences[@eng2015deeper, @yu2016pipi, @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast, @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath, @demichev2020dia] are counted and aggregated to obtain precursor abundances. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

Protein quantification proteomics experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with increasing numbers of samples are studied, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a specific factors and their interactions are significant. Furthermore it can be tested if differences of protein abundance between groups are significant. For this application it is important that the model provides an unbiased estimate of the false discovery rate (FDR) so that protein lists can be selected for follow-up experiments [TODO reference].   
Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins.

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analisis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level fold change and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage known from `r BiocStyle::Biocpkg("limma")`. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

When analyzing parallel-group designs using a single explanatory variable, and contrasting the groups, we can use all R packages; however, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that packages such as $limma$ and $proDA$ allow us to fit a comprehensive variety of models and test various hypotheses; however, in-depth knowledge of design matrix specification is required [@law2020guide].
